The role of cognateness in non-native spoken word recognition

### 2021-12-02 Rmd/manuscript.tex
Add comments by Chiara Santolin and APA 7th style

\hypertarget{native-speech-processing-is-effortless}{%
\subsection{Native speech processing is effortless}\label{native-speech-processing-is-effortless}}

Humans are able to recognise spoken words without much effort, even in adverse conditions such as compressed speech (De Haan, 1982) or the loss of segmental information (Warren, 1970). But even in the most ideal of the situations, the processes engaged in word recognition do not occur without uncertainty. The speech input activates multiple candidate lexical representations based on their similarity with the signal, and for word recognition to take place, one of the candidates must be selected. There is extensive literature about how the number of candidates, their frequency, and their phonological and semantic similarity with the target word impacts the dynamics of word recognition. But less is known about how these factors affect the recognition of words in a non-native language.

\hypertarget{non-antive-speech-processing-is-costly}{%
\subsection{Non-antive speech processing is costly}\label{non-antive-speech-processing-is-costly}}

Listening to speech in a non-native language is more costly than doing so in the native language, even if one is proficient in such language. The source of this increased cognitive effort is likely to originate at multiple levels. Some examples: (1) some sounds in the speech signal do not correspond to any phoneme in the native language, (2) segmenting the speech signal relies on previous familiarity with word forms or with the statistical regularities of the language: (3) word order might be different in both languages (e.g., subject-object-verb vs.~subject-verb-object.). The listener, however, is rarely completely naïve to the language they are listening to. All languages share, to some extent, similarities, frequently due to their typological closeness. These similarities which can be exploited by non-native listeners.

\hypertarget{cognateness-aids-non-native-speech-processing}{%
\subsection{Cognateness aids non-native speech processing}\label{cognateness-aids-non-native-speech-processing}}

One of such commonalities occurs at the lexical level, in the for of cognateness. Cognates are cross-language synonyms whose form (e.g., phonology, orthography, signature) is similar. The cause of this similarity is frequently attributed to a shared etymological origin. Romance languages like Spanish, Italian, French or Catalan, share many cognates (Schepens, Dijkstra, \& Grootjen, 2012). For example, this is the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and Catalan, respectively)\footnote{Some form-similar cross-language synonyms are technically not cognates. For example, \emph{sun} and \emph{sol} (in Spanish), share their phonological onset, but their etymology points to different origins. We will use the word \emph{cognateness} to include all form-similar cross-language synonyms for simplicity. It is highly implausible that etymology plays a direct role on language perception if it is not via form-similarity, since it is not necessary for participants in psycholinguistic experiments to be aware of the etymology of the words they encounter in the tasks to be subject to the effect of form-similarity.}.

\hypertarget{the-cognate-advantage-in-comprehension-production-and-learning}{%
\subsection{The cognate advantage in comprehension, production, and learning}\label{the-cognate-advantage-in-comprehension-production-and-learning}}

Cognates play a major role in virtually all models of bilingual lexical processing because they provide evidence that bilinguals access their lexicon in a language non-selective way: during word production and comprehension, both lexical representations are activated, and the form similarity between both impacts participants' performance in naming or lexical decision tasks (e.g., Costa, Caramazza, \& Sebastian-Galles, 2000; Thierry \& Wu, 2007). For instance, in their seminal study, Costa et al. (2000) asked Spanish-Catalan bilinguals to name pictures in Spanish. Unbeknownst to participants, half of the pictures' associated labels were cognates in Spanish and Catalan (e.g., \emph{puerta}-\emph{porta}) while the other half were non-cognates (e.g., \emph{mesa}-\emph{taula}). Surprisingly, participants named cognate pictures faster than non-cognate pictures, while Spanish monolinguals did not show this effect. This suggested that bilinguals activated picture's word representations in both languages, and that the phonological overlap between both labels facilitated the naming process.

There is also evidence that cognates are learnt more easily in L2 than non-cognates. For instance, De Groot and Keijzer (2000) presented 40 Dutch natives 60 pairs of translation equivalents. In each learning trial, two words were presented in a screen side-by-side: one in Dutch and one pseudoword. Pseudowords were generated in such way that were easily pronounceable and were phonotactically legal in Dutch. Word pairs varied in their form-similarity (number of shared letters, 40-70\%). Participants were tested twice in the same task with one week of difference. Participants' performance was better for cognates across both testing sessions, suggesting that cognates were learnt and retained more easily than non-cognates. Converging evidence was provided by Lotto and De Groot (1998) in Dutch natives learning Italian words: cognates were learnt more easily than non-cognates.

\hypertarget{the-cognate-advantage-in-translation-the-rmh-model}{%
\subsubsection{The cognate advantage in translation: the RMH model}\label{the-cognate-advantage-in-translation-the-rmh-model}}

Cognates are not only learnt faster than non-cognates, but translated faster too. The mechanisms behind this effect, though, are still unclear. Early accounts of bilingual lexical access suggested that low-proficiency learners first established links between newly learnt words in L2 and their meanings through their L1 translation equivalents (Potter, So, Von Eckardt, \& Feldman, 1984). As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on the mediation of L1 representations. The Revised Hierarchical Model {[}RHM; Kroll and Stewart (1994){]} captured this assumption and predicted that translating words from L1 to L2 (forward translation) should take longer than translating from L2 to L1 (backward translation). The rationale behind this prediction is that backward translation relies more strongly on direct word-word links between L1 and L2 representations, while forward translation would rely more strongly on the mediation between the concept and the two word forms. One of the consequences of this prediction is that backward translation should be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during backward translation.

To test these predictions, Degroot, Dannenburg, and Vanhell (1994) and Groot (1992) asked 52 Dutch natives with high (yet non-native) English proficiency to translate words from either Dutch to English (forward translation, L1 to L2) or from English to Dutch (backward translation, L2 to L1). In each trial, participants were presented visually with a word in English or Dutch, and were asked to speak out loud its translation in the other language as soon as possible. The authors reported three main findings: (1) translation times and accuracy were roughly equivalent in across both forward and backward conditions (although slightly faster in the former), (2) semantic variables, such as imageability, were positively associated with participants' performance more strongly in the forward translation conditional, although the effect size was small, (3) cognates were translated equally fast in both conditions, but non-cognates were translated faster during backward translation than during forward translation, and (4) when translating cognates participants' performance across both conditions was less sensitive to semantic variables than when translating non-cognates. These results prompted the authors to reject a hard version Kroll and Stewart (1994)'s account, and suggested that both conceptually mediated and direct translation routes are active during translation, but backward translation relies more strongly on direct links between L1 and L2 representations, which makes it more sensitive to cognateness than forward translation. When the authors tested a group of bilinguals with higher proficiency in English, their results pointed in the same direction. Subsequent studies did not find differences in participant's performance during forward and backward translation, or even found better performances in forward translation (Christoffels, De Groot, \& Kroll, 2006; Christoffels, Ganushchak, \& Koester, 2013), contrary to the predictions of the RHM.

\hypertarget{the-impact-of-neighbourhoods-lexical-processing-the-nam-and-the-biabia-models}{%
\subsection{The impact of neighbourhoods lexical processing: the NAM and the BIA/BIA+ models}\label{the-impact-of-neighbourhoods-lexical-processing-the-nam-and-the-biabia-models}}

The idealised lexicon considered by the models described above only considers word-word connections between translation equivalents. More recent studies have highlighted the role of the rich network of connections that given word establishes with other phonologically or conceptually related words. This notion was first introduced in monolingual research by Collins and Loftus (1975) in their theory of semantic processing, and later Luce and Pisoni (1998) formalised this dimension in their Neighbourhood Activation Model (NAM). In this model, lexical selection is mediated not only by the lexical frequency of the target word, and the number of activated candidates, but also by the frequency of the candidates. Luce and Pisoni (1998) designed a lexical decision task in which English native participants listened to a word across several conditions of noise-to-signal ratio. The authors then used a computerised lexicon to calculate the number of phonological neighbours around each of the presented words, based on phonetic similarity matrices, and the average lexical frequency of such neighbourhoods. Participants answered more faster and more accurately to high frequency words. Words from high density neighbourhoods were responded to more accurately, but more slowly. The average lexical frequency of the neighbourhood was associated to a decrease in accuracy. Low frequency words were responded to more accurately in low density neighbours than in high density neighbours. The authors concluded that, although both word lexical frequency and neighbourhood density can, in principle, facilitate spoken word recognition, this effect was modulated by the frequency of the neighbourhood: when phonological neighbours are more frequent, lexical selection is hindered (Goldinger, Luce, \& Pisoni, 1989; Luce, Pisoni, \& Goldinger, 1990).

The fact that phonological neighbourhoods and their structure impacts lexical processing poses an important question for bilingualism research. Are word representations in one language part of phonological or orthographic neighbours in the other language? Van Heuven, Dijkstra, and Grainger (1998) addressed this issue in an experimental series and proposed the Bilingual Interactive Activation (BIA, later revised and relabelled as BIA+). In line with connectionist accounts of lexical processing this model advocated for lexical representations to establish both excitatory and inhibitory connections with other representations from the same of the other language (McClelland \& Rumelhart, 1981), forming an integrated lexicon for both languages. Van Heuven et al. (1998) tested Dutch-English bilinguals in a progressive demasking task (in each trial a words in presented in increasingly longer time windows interrupted by a checkerboard until the participant presses a button to type the word in a keyboard) or a lexical decision task (words a visually presented and the participant must decide as quickly as possible whether its a word in the target language or a non-word). The authors manipulated by size of the orthographic neighbourhood size of the words presented in its language (target language) or in the other language (non-target language). Interestingly, the authors reported participants' performance to be affected by the neighbourhood size of the presented word in the non-target language across both tasks. Specifically, participants took longer to respond to words with high neighbourhood density in the non-target language, even when participants completed the task in the target language exclusively (e.g., were presented with Dutch words exclusively). Monolinguals, on the other hand, were only affected by neighbourhood size in the target language. This provided evidence of cross-language interference between the word representation in the non-target language and the those in the target language.

\hypertarget{is-the-lexical-route-necessary-for-translation-the-multilink-model}{%
\subsection{Is the lexical route necessary for translation?: The Multilink model}\label{is-the-lexical-route-necessary-for-translation-the-multilink-model}}

More recently, Dijkstra et al. (2019) implemented a localist-connectionist model, Multilink, that integrates and formalises previous claims and predictions on how an interactive account of bilingual lexical access impacts word recognition, production, and more relevant to the aims of the present study, translation. This model assumes (1) an integrated lexicon in which word representations from the two languages establish connections as words from the same language do; (2) during backward translation, the presented word in L2 can activate form-similar words in L1 that compete with the target word in L1; (3) contrary to the RHM model's assumption that translation equivalents are linked by excitatory connections (e.g., translation can be done via word-word connections), Dijkstra et al. (2019) suggest that such connections would spread activation though irrelevant words in the target language and therefore initially assumed that translation equivalents are exclusively connect throught their shared concept; (4) the strength of the association between L2 representations and their concept is a function of the language user's proficiceny in L2. Multilink considers the particular case of Dutch-English bilinguals to generated simulations and test them against experimental data. The authors simulated data from Multilink to mirror Christoffels et al. (2006) translation elicitation task, and reported a remarkable correlation between Multilink's simulations and participant's data, with forward translation being faster than backward translation, contrary to the RHM model predictions. More relevant to the present study is the fact that the model also generated data suggesting that cognates are translated faster than non-cognates, in line with previous literature on the cognate advantage during translation. Overall, evidence from this model provides evidence for an integrated bilingual lexicon and for a facilitatory effect of cross-language similarity during translation. One of the main conclusions of Dijkstra et al. (2019) study is that word-word connections are not necessary for translation, as suggested by Kroll and Stewart (1994): Multilink did not include such connections and still fitted experimental data. However, as they note, both their simulations and the data collected by Christoffels et al. (2006) corresponded to high-proficiency bilinguals. In the case of the model simulations, vocabulary sizes were balanced in both languages and word-concept connections were equally strong for each member of the translation pairs in both languages. This scenario is actually contemplated by the Kroll and Stewart (1994)'s RHM model. What Dijkstra et al. (2019) did not test, is the fit of Mutilink's predictions for low-proficiency, unbalanced bilinguals, whose vocabulary sizes are smaller or their word-concept connections are weaker in L2 than in L1. It is under this set of assumptions that the RHM model predicts participants to rely strongly on word-word connections to be able to translation words, especially from L2 to L1. This prediction, however, is untested in the Multilink model.

\hypertarget{the-present-study}{%
\subsection{The present study}\label{the-present-study}}

In this study, we explored the plausibility of the lexical route in backward translation by testing monolingual participants. Monolinguals can be considered a particular (extreme) case of unbalanced bilingualism, in which which participants' vocabulary size in L2 is null, and word-concept connections are absent (i.e., set to zero in formal terms). In this scenario, participants can only rely on the similarity between the L2 form (presented as stimulus) and the L1 form (the target word). In other words, these participants can only use the lexical route to suceeed in the translation task. If participants are able to translate words in L2, even with no knowledge of the language it belongs to, based on their form-similarity with the target word, this would sugges that the lexical route is in place for low-proficiency bilinguals. If participants are not able to translation words from L2 to L1 regardless of their form-similarity, this would suggest that low-proficiency bilinguals rely entirely on word-concept connections for backward translation.

@@

The experiment was implemented online using Psychopy/Pavlovia (Peirce et al., 2019). Participants accessed the study from a link provided by Prolific or SONA and completed the experiment from an internet browser (Chrome or Mozilla). After giving their consent for participating, participants answered a series of questions about their demographic status, their language background, and the set up they were using for completing the study. Then, participants completed the experimental task. Participants were informed that they would listen to a series of pre-recorded words in Catalan or Spanish (English participants) or only Catalan (Spanish participants). They were instructed to listen to each word, guess its meaning in English (English participants) or Spanish (Spanish participants), and type their answer as soon as possible. English participants were randomly assigned to the list of Catalan or Spanish trials. Participants in the Catalan list were presented with 86 trials, and participants in the Spanish list were presented with 103 trials.

\begin{figure}
\includegraphics[width=0.8\linewidth]{C:/Users/u155880/Documents/translation-elicitation/Figures/design} \caption{ }\label{fig:procedurefigure}
\end{figure}

Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ``\textgreater{}'' symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN key to start and new trial. We excluded trials where participants did not type an existing word in the correspondent language, or did not type anything at all. Trials where the response was mistyped by only one character were counted as correct, as long as the respond did not correspond to a distinct word. Participants contributed a total of 8077 valid trials (5235 in Catalan, 2842 in Spanish). The task took approximately 15 minutes to be completed.

@@

Finally, we measured the phonological similarity between translation pairs by computing the Levenshtein similarity between their IPA translations using the \texttt{stringsim} function of the stringdist R package (van der Loo, 2014). This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme symbols) by counting the number of additions, deletions, and substitutions necessary to make both strings identical (Levenshtein \& others, 1966). This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a score that ranges from 0 to 1, where 0 indicates no similarity between the two strings and one indicates that both strings are identical. We computed this similarity measure (Levenshtein, from now on) for every translation pair in our stimuli lists. Table 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@

\leavevmode\vadjust pre{\hypertarget{ref-collins1975spreading}{}}%
Collins, A. M., \& Loftus, E. F. (1975). A spreading-activation theory of semantic processing. \emph{Psychological Review}, \emph{82}(6), 407.

>>>>

Humans are able to recognise spoken words without much effort, even in adverse conditions such as compressed speech (De Haan, 1982) or the loss of segmental information (Warren, 1970). But even in the most ideal of the situations, the cognitive processes engaged in word recognition do not occur without uncertainty. The speech input activates multiple candidate lexical representations based on their similarity with the signal, and for word recognition to take place, one of the candidates must be selected. There is extensive literature about the influence on word recognition of the number of lexical candidates, their frequency, and their phonological and semantic similarity with target words. Less is known about how all these factors affect the dynamics of word recognition in non-native languages.

Listening to speech in a non-native language is more costly than doing so in native languages, regardless of the level of proficiency {[}REF{]}. The source of this increased cognitive effort is likely to originate at multiple levels. For example, some sounds in the speech signal do not correspond to any phoneme in the native language, making the process of matching sounds to phonemes complicated. This is the case of a Spanish native listening to words containing the guttural /ʁ/ in French. Language learners, however, are rarely completely naïve to the language they are listening to. All languages share, to some extent, similarities, frequently due to their typological closeness. These similarities which can be exploited by non-native listeners to process the speech signal. In the present research we investigate the extent to which adult listeners rely on phonological similarity between cross-language synonyms during word translation.

Cognates are cross-language synonyms whose form (e.g., phonology, orthography, signature) is similar. The cause of such similarity is frequently attributed to a shared etymological origin. Romance languages such as Spanish and Catalan share many cognates (Schepens, Dijkstra, \& Grootjen, 2012), as in the case of \emph{puerta} and \emph{porta} (\emph{door} in Spanish and Catalan, respectively)\footnote{Some form-similar cross-language synonyms are technically not cognates. For example, \emph{sun} and \emph{sol} (in Spanish), share their phonological onset, but their etymology points to different origins. We will use the word \emph{cognateness} to include all form-similar cross-language synonyms for simplicity. It is highly implausible that etymology plays a direct role on language perception if it is not via form-similarity, since it is not necessary for participants in psycholinguistic experiments to be aware of the etymology of the words they encounter in the tasks to be subject to the effect of form-similarity.}.

Cognates play a major role in virtually all models of bilingual lexical processing because they provide evidence that bilinguals access the lexicon in a language non-selective way. During word production and comprehension, lexical representations of both languages are activated, and the form similarity between both impacts bilinguals' performance in naming or lexical decision tasks (e.g., Costa, Caramazza, \& Sebastian-Galles, 2000; Thierry \& Wu, 2007). In a seminal study, Costa et al. (2000) asked Spanish-Catalan adult bilinguals to name pictures in Spanish. Unbeknownst to participants, half of the labels associated to pictures were cognates in Spanish and Catalan (e.g., \emph{puerta}-\emph{porta}) while the other half were non-cognates (e.g., \emph{mesa}-\emph{taula}, meaning \emph{table} in Spanish and Catalan). Surprisingly, Spanish-Catalan bilinguals named cognate pictures faster than non-cognate pictures, while Spanish monolinguals did not show this effect. This suggests that bilinguals activate the picture's word representations in both languages, and the phonological overlap between both labels facilitated the naming process.

Additionally, cognates seem to be learnt more easily in L2 than non-cognates. De Groot and Keijzer (2000) presented Dutch natives pairs of Dutch words and pseudowords. Participants' performance learnt and retained better pairs with high orthographic similarity than pairs with low orthographic similarity (see also Lotto \& De Groot, 1998 for Dutch natives learning Italian words).

Cognates are not only learnt faster than non-cognates, but translated faster too. Early accounts of bilingual lexical access suggest that low-proficiency learners first established links between newly acquired words in L2 and their meanings through their L1 translation equivalents (Potter, So, Von Eckardt, \& Feldman, 1984). As learners become more proficient, the connection between L2 representations and their meaning grows stronger, and L2 word processing becomes less reliant on the mediation of L1 representations. The Revised Hierarchical Model {[}RHM; Kroll and Stewart (1994){]} captured this assumption and predicted that translating words from L1 to L2 (forward translation) should take longer than translating from L2 to L1 (backward translation). The rationale behind this prediction is that backward translation relies more strongly on direct word-word links between L1 and L2 representations, while forward translation would rely more strongly on the mediation between the concept and the two word forms. One of the consequences of this prediction is that backward translation should be more sensitive to the form similarity between the L1 and the L2 representations: cognate words should be retrieved faster than non-cognate words during backward translation.

To test these predictions, Degroot, Dannenburg, and Vanhell (1994) and Groot (1992) asked Dutch natives with high (yet non-native) English proficiency to translate words from either Dutch to English (forward translation, L1 to L2) or from English to Dutch (backward translation, L2 to L1). Participants were presented with written words in English or Dutch, and asked to speak out loud their translation in the other language as fast as possible. Results who that translation times and accuracy were roughly equivalent in across both forward and backward conditions, and that semantic variables (e.g., concreteness), were positively associated with participants' performance, especially in the forward translation condition. Cognates were translated equally fast in both conditions, but non-cognates were translated faster during backward than forward translation, and translation of cognates was less affected by semantic variables than non-cognates.

These findings suggest that both conceptually mediated and direct translation routes are active during translation, although backward translation relies more strongly on direct links between L1 and L2 representations, which makes it more sensitive to cognateness (than forward translation). Interestingly, higher-proficiency bilinguals tested in the same task showed similar results. However, subsequent studies did not find differences in participant's performance during forward and backward translation, or even found better performances in forward translation (Christoffels, De Groot, \& Kroll, 2006; Christoffels, Ganushchak, \& Koester, 2013), contrary to the predictions of the RHM. Further models build upon this model, introducing a wider range of connections in their theorised lexica.

Since Kroll and Stewart (1994) proposed their RHM model, subsequent studies brought attention to the role of the network of connections that words establish with each other at the form-level (phonological or orthographic) or at the conceptual level. The Neighbourhood Activation Model {[}NAM; Luce and Pisoni (1998){]} suggests that lexical selection is mediated not only by the lexical frequency of the target word (and the number of activated candidates), but also by the frequency of the candidates.

Subsequent behavioural experiments (e.g., Luce \& Pisoni, 1998) show that English adult participants were tested in a lexical decision task in which words were presented across several conditions of noise-to-signal ratio. A computerised lexicon calculated number and average lexical frequency of phonological neighbours around each of the presented words. Participants answered more faster and more accurately to high frequency words, and words from high density neighbourhoods were also responded detected more accurately, but slowly. Therefore, average lexical frequency of neighbourhoods was associated to a decrease in accuracy. Low frequency words were in fact detected more accurately in low density than in high density neighbours. These findings suggested that both word lexical frequency and neighbourhood density can facilitate spoken word recognition, but this effect is modulated by the frequency of the neighbourhood; when phonological neighbours are more frequent, lexical selection is hindered (Goldinger, Luce, \& Pisoni, 1989; Luce, Pisoni, \& Goldinger, 1990).

The influence of phonological neighbourhoods and their structure on lexical processing poses rise important question in the field of bilingualism research. Do word representations in one language form part of the phonological or orthographic neighbours of the other language? The Bilingual Interactive Activation {[}BIA+; Van Heuven, Dijkstra, and Grainger (1998){]} proposes that lexical representations establish both excitatory and inhibitory connections with other representations from the same or the other language, forming an integrated lexicon for both languages (McClelland \& Rumelhart, 1981). Evidence supporting such integrated lexicon is provided by the fact that bilinguals' performance in word recognition tasks is sensitive to the orthographic neighbourhood density of the translation of the presented word. Van Heuven et al. (1998) presented bilingual participants with words in only one language (target language) in a lexical decision task. When the translation of the presented word in the non-target language was part of a large neighbourhood, reaction times were slower. This provides evidence that the presented words activated orthographic neighbourhoods in the non-target language, which slowed down word recognition.

A more recent model of bilingual lexical processing is Multilink Dijkstra et al. (2019), which integrates and formalises previous claims and predictions on how an interactive account of bilingual lexical access impacts word recognition, production, and more relevant to the aims of the present study, translation. This model assumes an integrated lexicon in which word representations from the two languages establish connections in a way that is similar to what words of a single language do. Such model also assumes that during backward translation, the presented word in L2 can activate form-similar words in L1 that compete with the target word in L1 and it proposes that translation equivalents are exclusively connected through their shared concept, contrary to the RHM model (which suggests that translation equivalents are connected by excitatory connections at the lexical level; e.g., translation can be done via word-word connections). In addition, this model suggests that the strength of the association between L2 representations and their concept is a function of language users' proficiency in L2. A remarkable correlation between Multilink simulations of Christoffels et al. (2006) translation elicitation task and behavioural data with Dutch-English bilinguals has been reported, showing forward translation being faster than backward translation, contrary to the predictions made by the RHM model. Importantly, the model also generates data supporting the claim that cognates are translated faster than non-cognates, in line with previous literature on the cognate advantage during translation. Overall, results from Multilink from this model are compatible with an integrated bilingual lexicon and with the existence of a facilitatory effect of cross-language similarity during translation.

The fact that Multilink fitted the experimental data from Christoffels et al. (2006) with remarkable success, even in absence of word-word connections between translation equivalents might put into question the necessity of such connections during translation, in favour of a simpler model in which backward translation relies exclusively on the association between L2 and L1 word forms via their shared concept. This claim, however, should be taken. It should be taken into account, however, that Christoffels et al. (2006) only tested fairly balanced bilinguals (i.e., bilinguals with high proficiency in L2). Additionally, data generated from Multilink model assumed similar vocabulary sizes across languages and, perhaps more critically in this regard, that word-concept connections were equally strong for L1 and L2 representations. These conclusions should therefore be taken with caution when extended to unbalanced bilinguals. It is low-proficiency bilinguals whom the RHM model predicts strong reliance on word-word connections during translation, especially during backward translation. This prediction remains untested my Multilink.

The role of this study was to investigate the plausibility of the lexical route as a mechanism exploited by monolingual adults during backward translation. Monolinguals can be considered a particular (extreme) case of unbalanced bilingualism, in which their vocabulary size in L2 is null, and word-concept connections are absent. In this case, monolinguals can only use the lexical route to suceeed in the translation task, and they can only do so by exploiting the cognateness of the words presented and their translation in the native language. If participants are able to translate words in L2, even with no knowledge of the language it belongs to, based on their form-similarity with the target word, this would suggest that the lexical route is in place for low-proficiency bilinguals. If participants are not able to translation words from L2 to L1 regardless of their form-similarity, this would suggest that low-proficiency bilinguals rely entirely on word-concept connections for backward translation.

@@

The experiment was implemented online using Psychopy/Pavlovia (Peirce et al., 2019). Participants accessed the study from a link provided by Prolific or SONA and completed the experiment from an internet browser (Chrome or Mozilla). After giving their consent for participating, participants answered a series of questions about their demographic status, their language background, and the set up they were using for completing the study. Then, participants completed the experimental task. Participants were informed that they would listen to a series of pre-recorded words in Catalan or Spanish (English participants) or only Catalan (Spanish participants). They were instructed to listen to each word, guess its meaning in English (English participants) or Spanish (Spanish participants), and type their answer as soon as possible. English participants were randomly assigned to the list of Catalan or Spanish trials. Participants in the Catalan list were presented with 86 trials, and participants in the Spanish list were presented with 103 trials. Each trial started with a yellow fixation point presented during one second on the centre of the screen over a black background. After one second, the audio started playing while the dot remained being displayed until the audio ended. Upon the offset of the fixation point and audio, participants were prompted to write their answer by a ``\textgreater{}'' symbol. Typed letters were displayed in the screen in real time to provide visual feed-back to participants. Participants were allowed to correct their answer. Then, participants pressed the RETURN key to start and new trial. We excluded trials where participants did not type an existing word in the correspondent language, or did not type anything at all. Trials where the response was mistyped by only one character were counted as correct, as long as the respond did not correspond to a distinct word. Participants contributed a total of 8077 valid trials (5235 in Catalan, 2842 in Spanish). The task took approximately 15 minutes to be completed.

\begin{figure}
\includegraphics[width=0.8\linewidth]{C:/Users/u155880/Documents/translation-elicitation/Figures/design} \caption{Schematic representation of a trial in the experimental task.}\label{fig:procedurefigure}
\end{figure}

@@

We measured the phonological similarity between translation pairs by computing the Levenshtein similarity between their IPA translations using the \texttt{stringsim} function of the stringdist R package (van der Loo, 2014). This function computes the inverse of the Levenshtein distance between two character strings as a proportion. First, it computes the edit distance between two character strings (in this case, phoneme symbols) by counting the number of additions, deletions, and substitutions necessary to make both strings identical (Levenshtein \& others, 1966). This measure is then divided by the maximum distance (according to the length of the longest string) and then subtracted from 1. The result is a score that ranges from 0 to 1, where 0 indicates no similarity between the two strings and one indicates that both strings are identical. We computed this similarity measure (Levenshtein, from now on) for every translation pair in our stimuli lists. Table 1 summarises the lexical frequency, phonological neighbourhood density and phonological overlap of the words included in the Catalan and the Spanish lists.

@@
